%Acknowledgments

\renewcommand{\baselinestretch}{2}
\small\normalsize
\hbox{\ }
 
\vspace{.5in}

\begin{center}
\large{Acknowledgments} 
\end{center} 

\vspace{1ex}

% I owe my gratitude to all the people who have made this thesis possible and because of whom my graduate experience has been one that I will cherish forever.

% First and foremost I'd like to thank my advisor, Professor Rajarshi Roy for giving me an invaluable opportunity to work on challenging and extremely interesting projects over the past four years. He has always made himself available for help and advice and there has never been an occasion when I've knocked on his door and he hasn't given me time. It has been a pleasure to work with and learn from such an extraordinary individual.

% I would also like to thank my co-advisor, Dr. Parvez Guzdar. Without his extraordinary theoretical ideas and computational expertise, this thesis would have been a distant dream. Thanks are due to Professor Robert Gammon, Professor Edward Ott and Professor Thomas Antonsen for agreeing to serve on my thesis committee and for sparing their invaluable time reviewing the manuscript.

% My colleagues at the nonlinear optics laboratory have enriched my graduate life in many ways and deserve a special mention. David DeShazer helped me start-off by rewriting the basic simulation code in a user-friendly format. Christian Silva provided help by setting up the GRENOUILLE apparatus and performing some of the simulations. My interaction with  Rohit Tripathi, Ryan McAllister, Vasily Dronov, Min-Young Kim, Elizabeth Rogers, William Ray, Jordi Garcia Ojalvo, Riccardo Meucci, Atsushi Uchida, and Fabian Rogister has been very fruitful. I'd also like to thank Wing-Shun Lam and Benjamin Zeff for providing the LaTex style files for writing this thesis.

% I would also like to acknowledge help and support from some of the staff members. Donald Martin's technical help is highly appreciated, as is the computer hardware support from Edward Condon, LaTex and software help from Dorothea Brosius and purchasing help from Nancy Boone.

% I owe my deepest thanks to my family - my mother and father who have always stood by me and guided me through my career, and have pulled me through against impossible odds at times. Words cannot express the gratitude I owe them. I would also like to thank Dr. Mohan Advani, Dr. Vasudeo Paralikar and Dr. Vinod Chaugule who are like family members to me.

% My housemates at my place of residence have been a crucial factor in my finishing smoothly. I'd like to express my gratitude to Sivasankar Pandeti, Jayakumar Patil, Amit Trehan and Punyaslok Purakayastha for their friendship and support.

% I would like to acknowledge financial support from the Office of Naval Research (ONR), Physics, for all the projects discussed herein.

% It is impossible to remember all, and I apologize to those I've inadvertently left out.

% Lastly, thank you all and thank God!

Great Lord of Life and wisdom! In Thy Name.
Which to transcend no flight of thought may claim.

Here, I would like to thank all the people who helped me throughout graduate school and made this dissertation possible.

%First of all, I greatly thank my advisor, Prof. Rob Patro. I believe working with Rob in graduate school will always be one of the biggest honors of my life. He provided a very professional, yet friendly environment in Combine-Lab that lets anyone grow as much as they desire. Not only is he a very knowledgeable scientist who is always willing to share his wisdom, he is also one of the most ethical people I have ever known.
First, I would like to express my gratitude towards my advisor, Prof. Rob Patro. Working with Rob throughout my journey has been one of the best experiences of my life. He fostered a professional, yet friendly environment in the Combine-Lab that allows members to grow greatly. Not only is he a very knowledgeable scientist who is always willing to share his wisdom but also he is one of the most virtuous and honorable people I have ever met. Thank you for letting me work alongisde you.

I am tremendously grateful to the respected committee members Mihai Pop, Michael Cummings, Erin Molloy, and John Dickerson at the University of Maryland for their insightful comments to improve my dissertation.

I also appreciate my senior collaborators, Michael Love, Charlotte Soneson, Carl Kingsford as well as my internship supervisor Guillaume Marçais.
I gained a lot of great experience by working alongisde them.

I would also like to thank all the members of Combine-Lab, especially the founding members, Hirak, Avi, Laraib, and Fatemeh who I learned a lot from throughout the years. None of the works in this dissertation would have been possible without their kind collaborations. I am also thankful to all the current members of Combine-Lab, Jamshed, Jason, Dongze, Noor, Skylar, and Hyeon Jin who always inspired me with their ideas and suggestions to my work.

I would like to thank all the friendly members of CBCB who created a great environment for me to do research. I greatly thank all the members of the Computer Science department both in University of Maryland (specially Barbara and Tom) and  Stony Brook University (specially Cindy and Kathy) where I started my graduate school.

I thank my friends back home in Iran who always supported me from thousands of miles away. I am also greatly  thankful to my friends in the US who made a very kind and warm environment in a foreign land for me.
Especially the friends I met recently who greatly eased the last steps for me.

Last, but not least, I am thankful to my family, Mohammad Sadegh, Zahra, Pooya, and Ghazal, who are very dear to me and I will always owe a lot to them. I am very grateful for my family and I hope I can always make them happy.